Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance in zooplankton

### 2022-06-12 text/Manuscript.Rmd
MS and bib edits

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. In metamorphosis, a life history strategy widely found in metazoans, downward swimming towards the seafloor is known to be associated with larval settlement and the initiation of metamorphosis, a dramatic body change from the larval stage, and UV-avoidance behaviour allows dispersal and the acquisition of different ecological niches. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. On the other hand, the role of NO in zooplankton has been reported in many animals, including sponges, cnidarians, molluscs, arthropods, annelids, echinoderms and ascidians, as a molecule that controls settlement and metamorphosis initiation. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Verasztó2018]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences [need to add tree]. To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in four cells in the apical organ region (Fig. 1A). The same four cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and axonal projections of the four *NOS*-expressing cells allowed us to identify the same cells as four interneurons in the ciliary-type photoreceptor cells (cPRCs) circuit from Platynereis reconstruction data (Fig. 1B). We named these NOS interneurons, INNOS (INNOSdl, dr, vl, vr) (Fig. 1C). The cPRC circuit has been reported as a neural circuit involved in c-opsin1 mediated UV avoidance behaviour (Verasztó et al., 2017). INNOS were presynaptic from the cPRCs and postsynaptic to the RGW interneurons in the cPRC circuit (Fig. 1D). To determine whether NO is involved in UV avoidance behaviour, we generated mutant lines with the CRISPR/Cas-9 system and examined their behavioural responses to UV light.

@@

**NOS mutant delay from UV avoidance**

We created two mutant lines with different deletions in the third exon of the NOS gene (NOSΔ11/Δ11 and Δ23/Δ23) (Fig. 2A). Compared to wild-type larvae, homozygous NOS mutant larvae have a defective UV-avoidance response (Fig. 2B). Whereas wild-type larvae showed downward swimming behaviour within 30 seconds in response to lateral UV light (UV-avoidance), the NOS mutant larvae showed a delayed onset of this behaviour and a reduced downward swimming speed (Fig. 2B). These results are similar to previously reported responses of UV-avoidance behaviour in c-opsin1 mutants (Verasztó et al., 2017), suggesting that NO is involved in UV-avoidance behaviour via the cPRC circuit. We therefore searched for target molecules for NO.

@@

**GCs with NIT (nitrite/nitrate sensing) domains (NIT-GCs) are expressed in cPRC**

@@

To determine the function of NO in the cPRC circuit, the intracellular dynamics of neurons in the cPRC circuit were investigated, focusing on NIT-GC1. First, we investigated the changes in NO during UV stimulation of the cPRC using DAF-FM, a fluorescent probe that can detect NO, in order to determine whether NO is produced by UV light. By injecting DAF-FM directly into zygotes, INNOS cells could be identified under live imaging. To visualise the location of cPRCs, the calcium indicator RGECO1a was also injected together, as cPRCs maintain a higher resting calcium levels compared to other cells. Direct UV stimulation to the cPRCs increased the fluorescence intensity of DAF-FM compared to stimulation of non-cPRC region (Fig. 5A). Next, we did calcium imaging of cells in the cPRC circuit during UV stimulation. We locally stimulated the cPRC cilia in wild-type with a 405 nm laser (UV) and detected a slight increase followed by a strong decrease in calcium levels (hyperpolarisation) (Fig. 5B). Turning off the UV after or during the decrease caused a strong increase (depolarisation) and a return to the same calcium level as before (Fig. 5B). These responses were similar to those reported previously (Verasztó et al., 2017). We performed a similar stimulation using NOS mutants, which hyperpolarised during the UV stimulation, but no depolarising responses were observed when UV was stopped (Fig. 5C). Similarly, in NIT-GC1 knockdown larvae using the morpholinos, no depolarisation after the hyperpolarisation response were observed (Fig. 5D). Interestingly, no hyperpolarising response during UV stimulation was found to occur in NIT-GC2 knockdown larvae. And we compared the changes of the INNOS calcium levels identified by their position in relation to cPRCs during UV stimulation in wild-type and in NOS mutants. In both cases, a depolarising response was observed at the same time as the hyperpolarising response of the cPRCs (Fig. 5E). Furthermore, to perform calcium imaging of INRGWs in UV stimulation, INRGWs were identified by immunostaining with RGWa antibodies after calcium imaging (Fig. 5F). It has been reported that the INRGW cells form many large synapses with synaptic vesicles on two serotonergic ciliomotor neurons (Ser-h1) that are known to play a role in the regulation of ciliary beating (Verasztó et al., 2017) and the activity of the INRGW cells increased the negative correlation between the activity patterns of serotonergic neurons and the MC neuron (Williams et al., 2018). The results showed that depolarisation of INNOS occurs at the same time as hyperpolarisation of cPRC in the wild type. Identified INRGW cells were found to cause depolarisation at the same time as cPRC hyperpolarisation in the wild type (Fig. 5G). It was also found that in NOS mutants where cPRC hyperpolarisation does not occur, the INRGW depolarisation response also does not occur (Fig. 5H).

@@

cPRCs expressing UV-absorbing c-opsin1 have high resting calcium levels and show hyperpolarising and subsequent depolarising responses to UV light (405 nm) [@Verasztó2018]. In the present study, we have shown that the cells that strongly synapse with cPRCs are interneurons that specifically express NOS (INNOS). The fluorescence intensity of DAF-FM, which specifically detected INNOS, showed a transient tonic-like increase dependent on UV stimulation of the cPRCs. The graph shows a decrease in intensity after UV light is switched off, but the response of DAF-FM to NO is irreversible and this decrease is due to breaching by the relatively strong excitation light. Attempts to measure changes in the intensity of the calcium indicator RGECO1a simultaneously with DAF-FM imaging did not lead to detection of changes in fluorescence intensity compared to GCaMP6s and could only visualise the cPRCs. In calcium imaging using GCaMP6s, INNOS identified by the positional relationship between the cPRC and INNOS and the labelling of RYa depolarised in clear synchrony with the hyperpolarising response of the cPRCs. This tonic response may involve dopaminergic nerves based on single-cell data (Fig) [@puopolo2001][@zhang2007][@Randel2014a]. In the mammalian retina, it is known that signals from bipolar photoreceptor cells are transmitted to downstream NOS-expressing amacrine cells (NOAC), where NO is produced [@jacoby2018]. These suggest that UV stimulation of cPRCs causes the production of NO via a depolarising response of INNOS.

@@

Another type of interneuron that strongly synapses with the cPRC is the INRGWa, which synapses directly onto the ciliomotor neuron and is therefore considered to be a very important neuron in the control of larval behaviour [@Verasztó2018][@williams2017]. The present results suggest that the depolarising response of the cPRC due to NO feedback induces INRGW activation. RGWamide is also a neuropeptide that induces larval downward swimming quite strongly (unpublished? need to show data?), and the delayed downward swimming of NOS mutant larvae in response to UV stimulation may be responsible for the failure of INRGW activation to occur. In addition, in many marine invertebrates, NO has been reported to be involved in many aspects of settlement behaviour and metamorphosis induction [@leise2001; @bishop2003; @bishop2007; @pechenik2007; @castellano2014; @ueda2013; @ueda2014; @ueda2016; @yang2018; @zhu2020; @song2021]. The downward swimming behaviour of Platynereis larvae is the first step in the onset of settlement behaviour and metamorphosis. These suggest that downward swimming due to NO activation of INRGWs may trigger the onset of settlement behaviour and metamorphosis [@Conzelmann2013][@Williams2015]. These suggest that NO feedback signalling may act as a trigger for downward swimming via activation of INRGWs, which induces the initiation of settlement behaviour and metamorphosis.

>>>>

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. In metamorphosis, a life history strategy widely found in metazoans, downward swimming towards the seafloor is known to be associated with larval settlement and the initiation of metamorphosis, a dramatic body change from the larval stage, and UV-avoidance behaviour allows dispersal and the acquisition of different ecological niches. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. On the other hand, the role of NO in zooplankton has been reported in many animals, including sponges, cnidarians, molluscs, arthropods, annelids, echinoderms and ascidians, as a molecule that controls settlement and metamorphosis initiation. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Veraszto2018]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (#Kei - need to add tree). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in four cells in the apical organ region (Fig. 1A) (#Kei - 3d HCR and transgene?). The same four cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1D-E). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Fig. 1F).

@@

**Nitric oxide produced during UV/violet stimulation gates the output of the cPRC circuit**

The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test if NO is produced during UV/violet responses, we injected zygotes with the fluorescent NO-reporter DAF-FM and imaged two-day-old larvae during the local stimulation of the cPRC cilia by 405 nm light. Following light stimulation, we detected an increase in DAF-FM fluorescence in the region of the INNOS cells. (#Kei - are there time course data that could be shown?). We next analysed the dynamics of the cPRC circuit following light stimulation. As we have shown previously [@Veraszto2018], a 30-sec illumination of the cPRC ciliary region lead to a decrease in cPRC calcium levels, likely due to the hyperpolarisation of the cPRCs. After 30 min, calcium levels in cPRCs were raising again, reaching higher levels than at the start of the stimulus - a response that may involve depolarisation. This activation phase ocurres after the end of the stimulation and may be due to a slower neuroendocrine feedback [@Veraszto2018]. We also quantified in more detail the responses of the INNOS and INRGW interneurons under a UV/violet stimulus to the cPRCs. To unambiguously identify the cells from which we recorded calcium signals, we developed an on-slide staining method to directly fix, immunostain and image the same larva that we imaged previously. We used the cell-specific neuropeptide markers anti-RYamide (INNOS) and anti-RGWa (INRGW) to mark the interneurons. Based on the position of the nuclei, we could correlate live and fixed samples at a single-cell precision. INNOS cells showed increase calcium signals during the inhibition phase of the cPRCs. In contrast, INRGW cells were only active during the second activation phase of cPRCs.\
To determine whether NO is involved in the observed circuit dynamics, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 3A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. Calcium imaging in *NOS*-mutant larvae revealed perturbed cPRC responses and circuit activation. In cPRCs, we could detect the inhibition phase but this was not followed by activation and calcium levels remained low. We did not detect a difference in the response of the INNOS cells between wild-type and *NOS*-mutant larvae. In contrast, INRGW cells did not activate in *NOS*-mutant larvae, suggesting that they will not initiate a circuit output (to Ser-h1 cells and the ciliary band).

**Nitric oxide signalling mediates UV-avoidance behaviour** Our calcium imaging data suggest that *NOS*-mutant larvae will have defective UV-avoidance behaviour. To test this, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated from two lateral sides with xy nm violet light. As previously shown, wild-type larvae swim downward within 30 sec following UV/violet light stimulation. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3B). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae.

@@

**Two unconventional guanylyl cyclases (NIT-GC) are expressed in the cPRCs**

@@

To determine the function of NO in the cPRC circuit, the intracellular dynamics of neurons in the cPRC circuit were investigated, focusing on NIT-GC1. First, we investigated the changes in NO during UV stimulation of the cPRC using DAF-FM, a fluorescent probe that can detect NO, in order to determine whether NO is produced by UV light. By injecting DAF-FM directly into zygotes, INNOS cells could be identified under live imaging. To visualise the location of cPRCs, the calcium indicator RGECO1a was also injected together, as cPRCs maintain a higher resting calcium levels compared to other cells. Direct UV stimulation to the cPRCs increased the fluorescence intensity of DAF-FM compared to stimulation of non-cPRC region (Fig. 5A). Next, we did calcium imaging of cells in the cPRC circuit during UV stimulation. We locally stimulated the cPRC cilia in wild-type with a 405 nm laser (UV) and detected a slight increase followed by a strong decrease in calcium levels (hyperpolarisation) (Fig. 5B). Turning off the UV after or during the decrease caused a strong increase (depolarisation) and a return to the same calcium level as before (Fig. 5B). These responses were similar to those reported previously [@Veraszto2018]. We performed a similar stimulation using NOS mutants, which hyperpolarised during the UV stimulation, but no depolarising responses were observed when UV was stopped (Fig. 5C). Similarly, in NIT-GC1 knockdown larvae using the morpholinos, no depolarisation after the hyperpolarisation response were observed (Fig. 5D). Interestingly, no hyperpolarising response during UV stimulation was found to occur in NIT-GC2 knockdown larvae. And we compared the changes of the INNOS calcium levels identified by their position in relation to cPRCs during UV stimulation in wild-type and in NOS mutants. In both cases, a depolarising response was observed at the same time as the hyperpolarising response of the cPRCs (Fig. 5E). Furthermore, to perform calcium imaging of INRGWs in UV stimulation, INRGWs were identified by immunostaining with RGWa antibodies after calcium imaging (Fig. 5F). It has been reported that the INRGW cells form many large synapses with synaptic vesicles on two serotonergic ciliomotor neurons (Ser-h1) that are known to play a role in the regulation of ciliary beating and the activity of the INRGW cells increased the negative correlation between the activity patterns of serotonergic neurons and the MC neuron (Williams et al., 2018). The results showed that depolarisation of INNOS occurs at the same time as hyperpolarisation of cPRC in the wild type. Identified INRGW cells were found to cause depolarisation at the same time as cPRC hyperpolarisation in the wild type (Fig. 5G). It was also found that in NOS mutants where cPRC hyperpolarisation does not occur, the INRGW depolarisation response also does not occur (Fig. 5H).

@@

cPRCs expressing UV-absorbing c-opsin1 have high resting calcium levels and show hyperpolarising and subsequent depolarising responses to UV light (405 nm) [@Veraszto2018]. In the present study, we have shown that the cells that strongly synapse with cPRCs are interneurons that specifically express NOS (INNOS). The fluorescence intensity of DAF-FM, which specifically detected INNOS, showed a transient tonic-like increase dependent on UV stimulation of the cPRCs. The graph shows a decrease in intensity after UV light is switched off, but the response of DAF-FM to NO is irreversible and this decrease is due to breaching by the relatively strong excitation light. Attempts to measure changes in the intensity of the calcium indicator RGECO1a simultaneously with DAF-FM imaging did not lead to detection of changes in fluorescence intensity compared to GCaMP6s and could only visualise the cPRCs. In calcium imaging using GCaMP6s, INNOS identified by the positional relationship between the cPRC and INNOS and the labelling of RYa depolarised in clear synchrony with the hyperpolarising response of the cPRCs. This tonic response may involve dopaminergic nerves based on single-cell data (Fig) [@puopolo2001][@zhang2007][@Randel2014a]. In the mammalian retina, it is known that signals from bipolar photoreceptor cells are transmitted to downstream NOS-expressing amacrine cells (NOAC), where NO is produced [@jacoby2018]. These suggest that UV stimulation of cPRCs causes the production of NO via a depolarising response of INNOS.

@@

Another type of interneuron that strongly synapses with the cPRC is the INRGWa, which synapses directly onto the ciliomotor neuron and is therefore considered to be a very important neuron in the control of larval behaviour [@Veraszto2018][@williams2017]. The present results suggest that the depolarising response of the cPRC due to NO feedback induces INRGW activation. RGWamide is also a neuropeptide that induces larval downward swimming quite strongly (unpublished? need to show data?), and the delayed downward swimming of NOS mutant larvae in response to UV stimulation may be responsible for the failure of INRGW activation to occur. In addition, in many marine invertebrates, NO has been reported to be involved in many aspects of settlement behaviour and metamorphosis induction [@leise2001; @bishop2003; @bishop2007; @pechenik2007; @castellano2014; @ueda2013; @ueda2014; @ueda2016; @yang2018; @zhu2020; @song2021]. The downward swimming behaviour of Platynereis larvae is the first step in the onset of settlement behaviour and metamorphosis. These suggest that downward swimming due to NO activation of INRGWs may trigger the onset of settlement behaviour and metamorphosis [@Conzelmann2013][@Williams2015]. These suggest that NO feedback signalling may act as a trigger for downward swimming via activation of INRGWs, which induces the initiation of settlement behaviour and metamorphosis.
